Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-03-22 ms.Rnw
minor edits from ELI
AIM seminar

Advent of routine molecular surveillance.
With certain pathogens, with matching of time-scales, information on transmission is possible.
With sufficiently dense sampling, it is sometimes possible to reconstruct WAIFW.
More typically, one wishes to gain insight into the structure and parameterization of the transmission process.
This is accomplished by formalizing a transmission model or suite of transmission models, estimating parameters, and comparing their ability to explain the data.

@@

Problem of phylodynamics.
Factorization of problem into two subproblems.

@@

@@

If $\Phir_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Phirh_k\coloneq{\Phir_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Phir_t=\Phirh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.

@@

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,I-1,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
    -\gamma\,I\,v(t,S,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,I,R+1)
    -\omega\,R\,v(t,S,I,R).
  \end{split}
\end{equation*}
Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,E-1,I,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,E,I,R)\\
    &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
    -\sigma\,E\,v(t,S,E,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
    -\gamma\,I\,v(t,S,E,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
    -\omega\,R\,v(t,S,E,I,R),
  \end{split}
\end{equation*}
where $N=S+E+I+R$ is the total population size.
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
This model has KFE
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
    -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
    -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
    -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
    -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
    -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
    -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
    -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
    -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
  \end{split}
\end{equation*}

@@

\Cref{fig:example_models}D depicts a model of superspreading.
There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
    -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
    &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
    -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
    -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
    -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
    -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
    -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
    -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
  \end{split}
\end{equation*}

@@

\begin{equation*}
  \frac{\partial{v}}{\partial{t}}=
  \lambda\,(N-1)\,v(t,N-1)
  -\lambda\,N\,v(t,N)
  +\mu\,(N+1)\,v(t,N+1)
  -\mu\,N\,v(t,N)
\end{equation*}

@@

\AAK{[Maybe it would be better to speak instead of $\Hr_t$ being the random $\sigma$-algebra generated by the sequences $(\Trh_k,\Xrh_k,\Urh_k)_{k=1}^{\Kr_t}$.
    The only thing we do with $\Hr$ is condition on it.
    One (perhaps minor) problem with this is that we cannot then speak about the measure on histories.
]}

@@

Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}
  \label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}=\Set{\Th_1,\dots,\Th_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.

@@

Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}).
Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
These occur whenever the color changes along a branch, but can also occur without a color-change.
We refer to such nodes as \emph{inline nodes}.

Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
In particular, let $\leaves$ be a countable set of labels and let $\func{Z}$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
That is, an element $\zeta\in{\func{Z}}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
Partition \emph{fineness} defines a partial order on $\func{Z}$.
Specifically, for $\zeta,\zeta'\in{\func{Z}}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\func{Z}$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
An element $b\in{Z_t}$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.
We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
Therefore, $\event{Z}\subseteq{\event{G}}$.

@@

The process just described yields a genealogy that relates all extant members of the population, and all samples.
Moreover, it details each lineage's complete history of movement through the various demes.
The data we ultimately wish to analyze will be based only on samples, however.
Nor, in general, will the histories of deme occupancy be observable.
A generative model must account for this loss of information.
We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches.

@@

Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(\eta,\eta')=1$ if and only if
there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
and a $t\in{[0,\T]}$ such that,
given $\Gr_\T=\G$ and $\Hr_\T=\H$,
$\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

\begin{lemma}
  \label{lemma:reg-filt}
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
  If $\Vr_t$ is an $\Rp$-valued random process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H}_t}{B(e,\Xt_e,\X_e)},
  \end{equation*}
  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$ that satisfies the filter equation
  \begin{equation}
    \label{eq:reg-filter-eq2}
    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
  \end{equation}
  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
\end{lemma}
\begin{proof}
  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
  Accordingly, as $\Delta\downarrow{0}$,
  \begin{equation*}
    \begin{aligned}
      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  In the limit, we obtain \cref{eq:reg-filter-eq}.
  The last statement follows by definition.
\end{proof}

\begin{lemma}
  \label{lemma:sing-filt}
  Suppose $\mathcal{M}\subset{\Rp}$ is countable such that $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$ is finite for all $t$.
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
  $\Vr_t$ is an $\Rp$-valued process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H}_{t}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H}_{t}}{B(e,\Xt_e,\X_e)}.
  \end{equation*}
  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$ that satisfies
  \begin{equation}
    \label{eq:two-part-filter}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)&=
      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      &t\notin{\mathcal{M}},\\
      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
      &t\in{\mathcal{M}}.
    \end{aligned}
  \end{equation}
  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
\end{lemma}

>>>>

With the advent of inexpensive genome sequencing technologies, routine molecular surveillance has become an important tool in the epidemiologist's kit.
When an infectious agent evolves quickly enough so that information on its history of transmission accumulates in its genome, yet not so quickly that it is dissipated, it is possible to extract this information to gain insight into the determinants of transmission.
With sufficiently dense sampling, it can be possible to reconstruct the specific history of who has infected whom, which can aid identification of the correlates of transmission and immunity and the features of infections.
More typically, one can attempt to learn about these features via the estimation of a model of the transmission process on the basis of relatively sparse samples.
In particular, one can formalize one or more mathematical models of transmission, estimate their parameters, and comparing their ability to explain data, following a standard statistical paradigm.
For the purposes of this paper, we refer to the latter program as \emph{phylodynamics}.

@@

Problem of phylodynamics.
Factorization of problem into two subproblems.
Problems with this approach \citep{Smith2017}.

@@

Modeling of the sampling process.

@@

If $\Phir_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Phirh^{}_k\coloneq{\Phir_{\Trh^{}_k}}$, $k=0,\dots,\Kr^{}_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Phir_t=\Phirh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.

@@

\citet{King2022} worked out formulas for the exact likelihood of a genealogy induced by an SIRS model.
The theory developed in this paper applies, but since there is only one deme in this model, this is a simple case.
Its state vector is $x=(S,I,R)$ and its KFE is
\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,I-1,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,I+1,R-1)
    -\gamma\,I\,v(t,S,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,I,R+1)
    -\omega\,R\,v(t,S,I,R).
  \end{split}
\end{equation*}
\normalfont
Note that we have here allowed for the possibility that the transmission rate, $\beta$, depends on time.

@@

A simple, yet interesting, model with more than one deme is the SEIRS model (\cref{fig:example_models}A).
The state space is $\Rp^4$, with the state $x=(S,E,I,R)$ defined by the numbers of hosts in each of the four compartments.
It has two demes ($\Demes=\Set{\lab{E},\lab{I}}$) and the KFE
\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}(t,S,E,I,R)=
    &\frac{\beta(t)\,I\,(S+1)}{N}\,v(t,S+1,E-1,I,R)
    -\frac{\beta(t)\,I\,S}{N}\,v(t,S,E,I,R)\\
    &+\sigma\,(E+1)\,v(t,S,E+1,I-1,R)
    -\sigma\,E\,v(t,S,E,I,R)\\
    &+\gamma\,(I+1)\,v(t,S,E,I+1,R-1)
    -\gamma\,I\,v(t,S,E,I,R)\\
    &+\omega\,(R+1)\,v(t,S-1,E,I,R+1)
    -\omega\,R\,v(t,S,E,I,R),
  \end{split}
\end{equation*}
\normalfont
where $N=S+E+I+R$ is the total population size.
The deme occupancy function in this case is $n(x)=(E,I)$.
Note that the terms associated with sampling cancel each other in the KFE, since, in this model, sampling has no effect on the state.

@@

A simple model for the competition of two strains for susceptible hosts is depicted in \cref{fig:example_models}B.
In this model, the state vector consists of seven numbers: $x=(S,E_1,E_2,I_1,I_2,R_1,R_2)$.
There are four demes ($\Demes=\Set{\lab{E}_1,\lab{E}_2,\lab{I}_1,\lab{I}_2}$) and the occupancy function is $n(x)=(E_1,E_2,I_1,I_2)$.
This model has KFE
\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta_1(t)\,I_1\,(S+1)}{N}\,v(t,S+1,E_1-1,E_2,I_1,I_2,R_1,R_2)
    -\frac{\beta_1(t)\,I_1\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\frac{\beta_2(t)\,I_2\,(S+1)}{N}\,v(t,S+1,E_1,E_2-1,I_1,I_2,R_1,R_2)
    -\frac{\beta_2(t)\,I_2\,S}{N}\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_1\,(E_1+1)\,v(t,S,E_1+1,E_2,I_1-1,I_2,R_1,R_2)
    -\sigma_1\,E_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\sigma_2\,(E_2+1)\,v(t,S,E_1,E_2+1,I_1,I_2-1,R_1,R_2)
    -\sigma_2\,E_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_1\,(I_1+1)\,v(t,S,E_1,E_2,I_1+1,I_2,R_1-1,R_2)
    -\gamma_1\,I_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\gamma_2\,(I_2+1)\,v(t,S,E_1,E_2,I_1,I_2+1,R_1,R_2-1)
    -\gamma_2\,I_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_1\,(R_1+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1+1,R_2)
    -\omega_1\,R_1\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
    &+\omega_2\,(R_2+1)\,v(t,S-1,E_1,E_2,I_1,I_2,R_1,R_2+1)
    -\omega_2\,R_2\,v(t,S,E_1,E_2,I_1,I_2,R_1,R_2)\\
  \end{split}
\end{equation*}
\normalfont

@@

\Cref{fig:example_models}D depicts a model of superspreading.
There are three demes ($\Demes=\Set{\lab{E},\lab{I_L},\lab{I_H}}$).
The state vector is $x=(S,E,I_L,I_H,R)$ and the KFE is
\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \begin{split}
    \frac{\partial{v}}{\partial{t}}=
    &\frac{\beta(t)\,(I_L+\theta\,I_H)\,(S+1)}{N}\,v(t,S+1,E-1,I_L,I_H,R)
    -\frac{\beta(t)\,(I_L+\theta\,I_H)\,S}{N}\,v(t,S,E,I_L,I_H,R)\\
    &\sigma\,(E+1)\,v(t,S,E+1,I_L-1,I_H,R)
    -\sigma\,E\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{LH}\,(I_L+1)\,v(t,S,E,I_L+1,I_H-1,R)
    -\eps_{LH}\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\eps_{HL}\,(I_H+1)\,v(t,S,E,I_L-1,I_H+1,R)
    -\eps_{HL}\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_L+1)\,v(t,S,E,I_L+1,I_H,R-1)
    -\gamma\,I_L\,v(t,S,E,I_L,I_H,R)\\
    &\gamma\,(I_H+1)\,v(t,S,E,I_L,I_H+1,R-1)
    -\gamma\,I_H\,v(t,S,E,I_L,I_H,R)\\
    &\omega\,(R+1)\,v(t,S-1,E,I_L,I_H,R+1)
    -\omega\,R\,v(t,S,E,I_L,I_H,R).\\
  \end{split}
\end{equation*}
\normalfont

@@

\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \frac{\partial{v}}{\partial{t}}=
  \lambda\,(N-1)\,v(t,N-1)
  -\lambda\,N\,v(t,N)
  +\mu\,(N+1)\,v(t,N+1)
  -\mu\,N\,v(t,N)
\end{equation*}
\normalfont

@@

@@

Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic, as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\fontsize{10pt}{11pt}\selectfont
\begin{equation*}
  \label{eq:Hdens}
  \pi^{\mathrm{H}}(\dd{\H_t})
  =p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
  \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}
  \,\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
\end{equation*}
\normalfont
where again, by convention, $\Th^{}_0=0$ and $\Th^{}_{K^{}_t+1}=t$.

If $\H$ is such a history, we define $\time(\H)$ to be the right endpoint of its domain and use the notation $\event{\H}\coloneq\Set{\Th^{}_1,\dots,\Th^{}_{\K_t}}\subset{[0,\time(\H)]}$ to denote the set of its jump times.

@@

Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted with a distinct color for each deme (\cref{fig:geneal}).
Note that a genealogy will in general have \emph{branch-point nodes}, \ie internal nodes with more than one descendant, but may also have internal nodes with only one descendant.
We refer to such nodes as \emph{inline nodes}.
These occur whenever the color changes along a branch, but can also occur without a color-change.

Formally, we define a genealogy, $G$, to be a triple, $(T,Z,Y)$, where $T=\time(G)\in\Rp$ is the \emph{genealogy time}, $Z$ specifies the genealogy's \emph{tree structure}, and $Y$ gives the \emph{coloring}.
In particular, let $\leaves$ be a countable set of labels and let $\part(\leaves)$ be the set of all collections of finite, mutually-disjoint subsets of $\leaves$.
That is, an element $\zeta\in{\part(\leaves)}$ is a partition of the finite set $\bigcup\zeta\subseteq\leaves$.
Partition \emph{fineness} defines a partial order on $\part(\leaves)$.
Specifically, for $\zeta,\zeta'\in{\part(\leaves)}$, we say $\zeta\preceq{\zeta'}$ if and only if for every $b'\in{\zeta'}$ there is $b\in{\zeta}$ such that $b\supseteq{b'}$.
The tree structure of $G$ is defined by a \cadlag\ map $Z:[0,T]\to\part(\leaves)$ that is monotone in the sense that $t_1\le{t_2}$ implies $Z_{t_1}\preceq{Z_{t_2}}$.
An element $b\in{Z_t}$ is a set of labels;
it represents the branch of the tree that bears the corresponding lineages.
We use the notation $\event{Z}$ to denote the set of times at which $Z$ is discontinuous.
Note that $\event{Z}$ includes the times of all tip, sample, and branch-point nodes, but excludes inline and root nodes.
Therefore, $\event{Z}\subseteq{\event{G}}$.

@@

The process just described yields a genealogy that relates all extant members of the population, and all samples.
Moreover, it details each lineage's complete history of movement through the various demes.
The data we ultimately wish to analyze will be based only on samples, however.
Nor, in general, will the histories of deme occupancy be observable.
A generative model must account for this loss of information.
We therefore now describe how genealogies are \emph{pruned} to yield sample-only genealogies and then \emph{obscured} via the erasure of color from their branches (\cref{fig:upo}).

@@

Suppose $P$ is a pruned genealogy, with $\time(P)=T$ and $t\in\event{P}$.
The local structure of $P$ at $t$ is, in general, compatible with only a subset of the possible jumps $\Jumps$.
For example, if the event in $P$ at $t$ is a branch node or a sample node, then it is compatible only with birth-type or sample-type jumps, respectively.
Similarly, if the node in $P$ at time $t$ is one at which a lineage moves from deme $i$ to deme $i'$, then $u$ must be either of $i\to{i'}$ migration type or of a birth type with parent in $i$ and $r^u_{i'}>0$.
To succinctly accommodate all possibilities, let us introduce the indicator function $Q$ such that $Q=1$ if the local genealogy structure---which is captured by the values of $P^{\lab{Y}}$ just before and after $t$---is compatible with an event of type $u$ and $Q=0$ otherwise.
That is, $Q_u(\eta,\eta')=1$ if and only if
there is a feasible genealogy, $\G=(\T,\Z,\Y)$, and history, $\H$,
and a $t\in{[0,\T]}$ such that,
given $\Gr_\T=\G$ and $\Hr_\T=\H$,
we have $\U_t=u$, $\Yt_t=\eta$, and $\Y_t=\eta'$.
We refer to $Q$ as the \emph{compatibility indicator}.

@@

\begin{lemma}
  \label{lemma:reg-filt}
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable.
  If $\Vr_t$ is an $\Rp$-valued random process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\prod_{e\in\event{\H_t}}{B(e,\Xt_e,\X_e)},
  \end{equation*}
  then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued deterministic process on $\Xspace$ with density $w(t,x)$ that satisfies the filter equation
  \begin{equation}
    \label{eq:reg-filter-eq2}
    \frac{\partial{w}}{\partial{t}}=\int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}}.
  \end{equation}
  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
\end{lemma}
\begin{proof}
  The expectation can be broken into three terms, according to whether $\H_t$ has zero, one, or more than one event in $\halfclosed{t-\Delta,t}$.
  Accordingly, as $\Delta\downarrow{0}$,
  \begin{equation*}
    \begin{aligned}
      w(t,x)=&\left(1-\Delta\,\int{\alpha(t-\Delta,x,x')\,\dd{x'}}\right)\,w(t-\Delta,x)\\
      &\qquad+\Delta\,\int{\alpha(t-\Delta,x',x)\,B(t-\Delta,x',x)\,w(t-\Delta,x')\,\dd{x'}}+o(\Delta).
    \end{aligned}
  \end{equation*}
  In the limit, we obtain \cref{eq:reg-filter-eq}.
  The last statement follows by definition.
\end{proof}

\begin{lemma}
  \label{lemma:sing-filt}
  Suppose $\mathcal{M}\subset{\Rp}$ is countable such that $\mathcal{M}_t\coloneq\mathcal{M}\cap{[0,t]}$ is finite for all $t$.
  Suppose that $B:\Rp\times\Xspace^2\to\Rp$ is measurable and
  $\Vr_t$ is an $\Rp$-valued process satisfying
  \begin{equation*}
    \CondExpect{\Vr_t}{\Hr_t=\H_t}=\Indicator{\event{\H_{t}}\supseteq\mathcal{M}_t}\,\prod_{e\in\event{\H_{t}}}{B(e,\Xt_e,\X_e)}.
  \end{equation*}
  Then $\CondExpect{\Vr_t}{\Xr_t=x}$ is a measure-valued process on $\Rp^{\vert\mathcal{M}_t\vert}\times\Xspace$ with density $w(t,x)$ that satisfies
  \begin{equation}
    \label{eq:two-part-filter}
    \begin{aligned}
      \frac{\partial{w}}{\partial{t}}(t,x)&=
      \int{w(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}}-\int{w(t,x)\,\alpha(t,x,x')\,\dd{x'}},\qquad
      &t\notin{\mathcal{M}},\\
      w(t,x)&=\int{\wt(t,x')\,\alpha(t,x',x)\,B(t,x',x)\,\dd{x'}},\qquad
      &t\in{\mathcal{M}}.
    \end{aligned}
  \end{equation}
  Moreover, $\Expect{\Vr_t}=\int{w(t,x)\,\dd{x}}$.
\end{lemma}
